<!-- README.md is generated from README.qmd via: quarto render README.qmd --to gfm -->

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  comment = "##",
  fig.path = "man/figures/README-",
  fig.width = 12,
  fig.height = 8,
  fig.align = "center",
  dev = "png"
)
```


# gamlss.prepdata: Preparing Data for Distributional Regression

## Overview

The purpose of this package is to provide functions to facilitate the creation
of `data.frame`'s suitable for  statistical modeling analysis and especially
distributional regression models using
[`gamlss`](https://cran.R-project.org/package=gamlss) and.
[`gamlss2`](https://github.com/gamlss-dev/gamlss2)  packages.

There is a lot of information which can be gain using a preliminary data
analysis. One could seeking information about the _variables_ in the data set
themselves, on _outliers_, on _associations_  between variables, what  _type_ of
relationships  exist between the response and the explanatory variables (linear
or not linear) and  possible _interactions_ between the explanatory variables. 
In addition, one, at this stage, could  decide  appropriate  _partitions_ of to
maximize statistical inference. While answers to those questions are not
necessarily final, at this pre-modeling stage, they could help at the next
stage of the fitting  process. All functions in the package can be used before
the functions  \texttt{gamlss()} or \texttt{gamlss2()} are used for modeling. 

## Installation

The package is not yet on CRAN but can be installed from
[R-universe](https://gamlss-dev.R-universe.dev/).

```{r}
#| label: install
#| eval: false
install.packages("gamlss.prepdata", repos = "https://gamlss-dev.R-universe.dev")
```

## Functions

The functions for manipulation of variables are shown below

| Functions                | Usage                                  |
|:-------------------------|:---------------------------------------|
| `data_dim()`             | Dimensions & % of omitted observations |
| `data_names()`           | Names of the data                      |
| `data_distrinct()`       | Distinct values in variables           |
| `data_which(()`          | NA's in variables                      |
| `data_str()`.            | The class of  variables etc.           |
| `data_omit()`            | Omit all the `NA`s                     |
| `data_char2fac()`        | From characters to factors             |
| `data_few2fac()`         | From few distinct obs. to factors      |
| `data_int2num()`         | From integers to  numeric              |
| `data _rm()`             | Remove variables                       |
| `data _rm1val()`         | Remove factors with one level          |
| `data _rename()`         | Rename variables                       |
| `data _renamove()`       | Remove variables                       |
| `data _select()`         | Select variables                       |
| `data _exclude_class()`  | Exclude a specified class              |
| `data _only_continous()` | Includes only continuous               |
| `data_rmNAvars()`        | Remove variables with NA values        |
| `data_fac2num()`         | Make factor numeric                    |

The function for graphics are show below

| Functions           | Usage                                                |
|:--------------------|:-----------------------------------------------------|
| `data_plot()`       | Univariate plots of all variables                    |
| `data_xyplot()`     | Pairwise plots of the response against all others    |
| `data_bucket()`     | Bucket plots of all variables                        |
| `data_cor(()`       | Pairwise correlations                                |
| `data_pcor(()`      | Pairwise partial correlations                        |
| `data_void()`       | Pairwise % of empty spaces                           |
| `data_inter()`      | Pairwise interactions                                |
| `data_response()`   | Response variable plots                              |
| `data_zscores()`    | Univariate plots using z-scores                      |
| `data_outliers()`   | Univariate detection of outliers                     |
| `data_leverage()`   | Detection of outliers in the x's  space              |
| `data_scale()`      | Univariate scaling the x's                           |
| `data_trans_plot()` | Checking for univivariate transformations in the x's |


## Graphics


Here some of the  graphical functions examples

```{r}
#| label: packages
#| fig-show: hide
library("gamlss")
library("gamlss2")
library("ggplot2")
library("gamlss.ggplots")
library("gamlss.prepdata")
library("dplyr") 
packageVersion("gamlss.prepdata")
da <- data_rm(rent99, c(2, 9)) 
dim(rent99)
dim(da)
```

## `data_plot()` {#sec-data_plot}

The function `data_plot` plots all the variable of the data individually. It plots;

-   the continuous variable as *histograms* with a density plots superimposed, (see the plots for `rent` and `yearc` below). Alternatively a *dot plots* can be requested, (see the example in @sec-data_xyplot).

-   the integers as *needle* plots, (see the plot for `area` below).

-   the categorical variables, as *bar* plots, (see the plots for `location`, `bath` `kitchen` and `cheating` below).

The message `100 % of data are saved` below is the result of the function `data_cut()` which is use before any `ggplot2` plot.

```{r}
#| label: data_plot
da |> data_plot()
```

The function could saves the `ggplot2` figures.


## `data_xyplot()` {#sec-data_xyplot}

The functions `data_xyplot()` plots the response variable against each of the independent explanatory variables. It plots the continuous against continuous as *scatter* plots and continuous variables against categorical as *box* plot.

::: callout-warning
At the moment there is no provision for categorical response variables.
:::

```{r}
#| label: data_xyplot
da |> data_xyplot(response=rent )
```

The output of the function saves the `ggplot2` figures.

## `data_bucket()` {#sec-data_bucket}

The function `data_bucket` can be used to identifies hight skewness and kurtosis on continuous variables in the data. Note that if the continuous variable is normaly distribueterd looking should be in the center of the figure.

```{r}
#| label: data_bucket
data_bucket(da, response=rent )
```
